---
title: "XXX"
author: "The IsoriX core team"
date: '`r Sys.Date()`'
output:
  pdf_document
vignette: >
  %\VignetteIndexEntry{XXX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, initialization, echo=FALSE, results='hide', message=FALSE}
library(knitr)
library(IsoriX)
library(raster)
# set global options for knitr (if changed, all cashed chunked will be rerun!)
knitr::opts_chunk$set(cache=TRUE, cache.path="./cache_knitr/", fig.path="./fig_knitr/", global.par=TRUE, width="\\linewidth", fig.align="center", dev="CairoPNG", dpi=500) # use dpi=100 for small pdfs
do.eval <- TRUE ## to check that it does work, put TRUE
```

Welcome to __IsoriX__, in this vignette we present the steps required for you to perform several usefull GIS related taks in IsoriX.

# Before starting

Please read the vignette ___Workflow___, if you haven't done so. You can access it by simply typing:

```{r, workflow, eval=FALSE}
vignette("Workflow", package="IsoriX")
```

We assume that you are still in the same R session than the one created during the simple workflow example. That implies that __IsoriX__ is loaded and that the objects we will use below have already been created.


```{r, load isoscape, echo=FALSE, message=FALSE, results='hide'}
data(countries)
data(oceanmask)
data(Europefit)
load("../vignette_workflow/isoscape.rda")
load("../vignette_workflow/assignment.rda")
elevationraster <- raster("../vignette_workflow/gmted2010_30mn.tif")
elev <- relevate(elevation.raster=elevationraster, isofit=Europefit, aggregation.factor=10)
```

You can test if you have all the object we need by running the following code:

```{r test if objects are here}
all(c("countries", "oceanmask", "elev", "isoscape", "assignment") %in% ls())
```

If the result is TRUE, you can proceed! If not, then go back to the vignette ___Workflow___.

## Pimping up your plots
### Add a 3D-image to your plot
```{r hillshade I}
slope <- terrain(x = elev, opt = "slope",unit = "radians", neighbors = 8)
aspect <- terrain(x = elev, opt = "aspect",unit = "radians", neighbors = 8)
hillsh <- hillShade(slope, aspect, angle = 45,direction = 270)
```

```{r hillshade II,warning=FALSE,fig.pos='t', fig.width=4, fig.height=3}
plot(hillsh, col=grey(0:100/100), legend=FALSE)
#and colors on top: alpha value gives semi-transparency
image(isoscape$isoscape$mean, col=terrain.colors(100,alpha=0.3),add=T)
points(isoscape$sp.points$sources,cex=0.5,pch='+')
plot(oceanmask,col='grey',border='transparent',add=T)
plot(countries,col='transparent',border='white',add=T)
box()
```


### Add Gridlines
```{r gridline, fig.pos='t', fig.width=3, fig.height=3}
plot(hillsh, col=grey(0:100/100), legend=FALSE)
#and colors on top: alpha value gives semi-transparency
image(isoscape$isoscape$mean, col=terrain.colors(100,alpha=0.3),add=T)
points(isoscape$sp.points$sources,cex=0.5,pch='+')
plot(oceanmask,col='grey',border='transparent',add=T)
plot(countries,col='transparent',border='white',add=T)
box()
grd <- gridlines(isoscape$isoscape$mean, 
                    easts = pretty(bbox(isoscape$isoscape$mean)[1,]),
                    norths = pretty(bbox(isoscape$isoscape$mean)[2,]), 
                    ndiscr = 20)
plot(grd, add=T, col='pink', lty=2)
```


### Add North Arrow and Scale Bar
```{r north arrow, fig.pos='t', fig.width=3, fig.height=3}
plot(hillsh, col=grey(0:100/100), legend=FALSE)
#and colors on top: alpha value gives semi-transparency
image(isoscape$isoscape$mean, col=terrain.colors(100,alpha=0.3),add=T)
points(isoscape$sp.points$sources,cex=0.5,pch='+')
plot(oceanmask,col='grey',border='transparent',add=T)
plot(countries,col='transparent',border='white',add=T)
box()
grd <- gridlines(isoscape$isoscape$mean, 
                    easts = pretty(bbox(isoscape$isoscape$mean)[1,]),
                    norths = pretty(bbox(isoscape$isoscape$mean)[2,]), 
                    ndiscr = 20)
plot(grd, add=T, col='pink', lty=2)
library(GISTools)
library(maps)
north.arrow(-20, 40, 2.25, cex.lab=0.7, col="grey80")
GISTools::map.scale(-15, 35, 10,"Km",1, 1000, sfcol='red')   #should be approx correct
```


### Change the projection of your map
For certain parts of the world it is better to use a different projection than our geographic coordinate system (WGS84). Data can easily be projected in R, however, there is no 'on the fly' projection like in GIS systems - each layer has to be projected separately. To find the optimal reference system for your area, please check here:
http://www.spatialreference.org
As an example, we will use a Mollenweide-projection. There is different commands for vector and raster data.

```{r projections}
sources_moll <-  spTransform(isoscape$sp.points$sources,   CRS("+proj=moll +datum=WGS84"))
wrld_moll <- spTransform(countries,   CRS("+proj=moll +datum=WGS84"))
ocma_moll <- spTransform(oceanmask,   CRS("+proj=moll +datum=WGS84")) ##### ALEX: the oceanmask get inverted, shiott...

iso_mean_moll <- projectRaster(isoscape$isoscape$mean, crs=CRS("+proj=moll +datum=WGS84"))
hillsh_moll <- projectRaster(hillsh, crs=CRS("+proj=moll +datum=WGS84"))
```


### Redo the plot

```{r hillshade IIb,warning=FALSE,fig.pos='t', fig.width=4, fig.height=3}
plot(hillsh_moll, col=grey(0:100/100), legend=FALSE)
image(iso_mean_moll, col=terrain.colors(100,alpha=0.3),add=T)
points(sources_moll,cex=0.5,pch='+')
#plot(ocma_moll,col='grey',border='transparent',add=T)############## taken out
plot(wrld_moll,col='transparent',border='white',add=T)
box()

```







