

The function `trellis.last.object()` is very handy: it retrieves the last plot created using __lattice__ and our plot functions for isoscapes and assignments use __lattice__ since they use __rasterVis__, which is a package itself based on __lattice__.

As you can see the location of origin (purple) and the location for the best assignment (orange) differ a bit. The flexibility of __IsoriX__ helps to explore potential sources of problems and limitation. We can for example compare our isoscape prediction in these two locations:

```{r comparison}
print(paste("mean isoscape value at origin (+/-SE) =",
            round(extract(isoscape$isoscapes$mean, origin), 2), "+/-",
            round(sqrt(extract(isoscape$isoscapes$mean.predVar, origin)), 2)))

print(paste("mean isoscape value at maximum (+/-SE) =",
            round(extract(isoscape$isoscapes$mean, maximum),2), "+/-",
            round(sqrt(extract(isoscape$isoscapes$mean.predVar, maximum)), 2)))
```

Note that the estimate for the mean isotopic value where the maximum is has a large standard error as this area is not well covered by sources. You can also observe this effect by simply looking at the map for the prediction isoscape variance which we have plotted above. In this context, it is thus expected that such area will be difficult to rule out as a possible source of origin. This example is a good reminder that one must always think about the quality of the isoscape before drawing definitive biological conclusion.

Another way to look at our assignment is to check if for all bats the predicted assignment region does include the real location of origin (which is possible because here we do know what the real origin of the bats is). To do so, we start by extracting the p-values of the assignment at the location of origin for all individuals:

```{r ground truthing}
pvs <- c(extract(assignment$indiv$pv, origin))
```

We can then count for how many bats the true origin location is not rejected by our assignment test:

```{r ground truthing 2}
table(pvs > 0.05)
```

All bats are in! That is great but it will not necessarily happen all the time, especially if as here the calibration is performed on a different species than the individuals you want to allocate. As an exercise you can now try to plot the bats we assigned into the calibration fit to see if they do behave in the same way as other species.
