---
title: "Coding style guide for IsoriX"
author: "The IsoriX core team"
date: "`r Sys.Date()`"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Coding style guide for IsoriX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Assignment

Assignments are done using an arrow ```<-```, not ```=```.
```{r message=FALSE, warning=FALSE, results="hide"}
## Example:
y <- 2
```

## Notation and naming

### Objects

Datasets should be named in the format `SomethingType`.

### Classes

Classes should have a name of the format `somethingVerb`.

### Functions

The name of internal functions should be of the format `dosomething_somethingelse`, where the input is `something_somethingelse` and the output for the global environment sould be `my_something`.

## Curly braces

The opening curly braces are seperated with one white space and the closing ones go on a new line and are aligned with the first command.
The else statement is surrounded by curly braces on the same line.

```{r message=FALSE, warning=FALSE, results="hide"}
## Example for an if-condition:
if (y == 0) {
  log(y)
} else {
  y ^ 2
}
```

The same goes for a function except we do not put a space before the left parantheses. 

```{r message=FALSE, warning=FALSE, results="hide"}
## Example for a function:
function(x) {
  x + x
}
```

## Organisation

### Indentation and Spacing

We have 2 spaces indentation throughout the whole code. An exception is made when a line break occurs inside parentheses, then the wrapped line is aligned with the first character inside the parenthesis.
```{r message=FALSE, warning=FALSE, results="hide", fig.keep = "none"}
## Example:
plot(x = sort(rnorm(47)), 
     type = "s", 
     main = "title")
```

Spaces go around all operators (```+```, ```=```, ```<-```, ```!=```, ```==```, etc.) for improved readability.

### Commenting guidelines

For comments inside the code we use a douple hash ```##``` to seperate real comments from commented code chunks with only a single hash ```#```.