# Building Isoscapes {#isoscape}
```{r setup_isoscape, echo=FALSE}
suppressPackageStartupMessages({
  library(IsoriX, warn.conflicts = FALSE)
  library(knitr, warn.conflicts = FALSE)
  library(kableExtra, warn.conflicts = FALSE)
})
options(width = 80)
opts_chunk$set(cache = FALSE,
               global.par = TRUE,
               fig.align = "center",
               error = TRUE)
```

## Preparing the raw data

To get an isoscape, you need data capturing how the isotope composition varies in space.
In this documentation, we will use data provided by the Global Networks of Isotopes in Precipitation (GNIP), but you can use whatever source of data you prefer, or have access to.
In any case, do make sure to format your data the same way we do it.
Precisely, your dataset should be a `data.frame` (or `tibble`) with the same columns as the ones shown below in section \@ref(GNIPDataDE).

### The GNIP database {#GNIP}

You can download precipitation isotope data for hydrogen and oxygen from the Global Networks of Isotopes in Precipitation (GNIP).

This step must be done outside IsoriX since there is no API for GNIP.

To get to know what the GNIP is, its history, and more importantly its terms of use and information about the data, go there: http://www-naweb.iaea.org/napc/ih/IHS_resources_gnip.html

The GNIP data are free to download after the registration process is completed.
The following link will bring you to the right place to create an account:
https://websso.iaea.org/IM/UserRegistrationPage.aspx?returnpage=http://nucleus.iaea.org/wiser

Once your account has been activated, download the data you need from here:
https://nucleus.iaea.org/wiser/index.aspx (Tab Datasets)

For the time being, downloads are limited to 5,000 records per batch, which makes the compilation of huge databases fastidious.
GNIP promised to come up, in the future, with datasets directly prepared for IsoriX to save you the trouble.
We are eagerly waiting for this to happen!

### A toy example: GNIPDataDE {#GNIPDataDE}

Within IsoriX we have already included a small extract from GNIP for you to try out the functions of the package without having to worry about the data.
This small dataset corresponds to $\delta^2H$ values for Germany.
The dataset has kindly been provided by Christine Stumpp.
Here is what the beginning of the dataset looks like:

```{r GNIPDataDE, eval = FALSE}
head(GNIPDataDE)
```

```{r GNIPDataDE_bis, echo = FALSE}
kable(head(GNIPDataDE)) %>% kable_styling(bootstrap_options = "striped", font_size = 15)
```

### A real example: GNIPDataEU {#GNIPDataEU}

Here we are going to show you how to prepare a dataset for IsoriX from a raw extract from GNIP.
For this, we will use a file in which we have compiled the GNIP data for $\delta^2H$ values for the whole world.
We are unfortunately not being able to share these data with you, but you can download these data yourself as explained in section \@ref(GNIP).

We first import the data into R:

```{r GNIP_raw}
rawGNIP <- read.csv("./data/2016-10-31 Extract_ISORIX.csv")
```

This dataset contains `r nrow(rawGNIP)` rows and `r ncol(rawGNIP)` columns.
For example, the first row contains the following information:

```{r GNIP_raw_1st_row, echo = FALSE}
## we display the first row for this dataset as a column for visualization
kable(t(rawGNIP[1, ])) %>% kable_styling(bootstrap_options = "striped", font_size = 15)
```

We are now going to reshape these data step-by-step to make them ready for IsoriX.
We first reformat some temporal information:

```{r GNIP_date_extraction}
rawGNIP$year.begin  <- as.numeric(format(as.Date(rawGNIP$Begin.Of.Period), "%Y"))
rawGNIP$year.end    <- as.numeric(format(as.Date(rawGNIP$End.of.Period), "%Y"))
rawGNIP$year.span   <- rawGNIP$year.begin - rawGNIP$year.end
rawGNIP$month.begin <- as.numeric(format(as.Date(rawGNIP$Begin.Of.Period), "%m"))
rawGNIP$month.end   <- as.numeric(format(as.Date(rawGNIP$End.of.Period), "%m"))
rawGNIP$day.span    <- as.Date(rawGNIP$Begin.Of.Period) - as.Date(rawGNIP$End.of.Period)
rawGNIP$Year        <- as.numeric(format(as.Date(rawGNIP$Date), "%Y"))
rawGNIP$Month       <- as.numeric(format(as.Date(rawGNIP$Date), "%m"))
```

Mind that since we downloaded that file WISER has renamed `Begin.Of.Period` to `Begin.of.Period` (small case for the "o" in `of`), so you may need to adjust that.
Do pay attention to other possible change that may have also occurred in WISER.

Second, we identify the rows for which crucial information is missing.
Because we are going to make an isoscape based on deuterium measurements, we only check for the availability of data for this isotope in particular.
If you work on oxygen, you will have to adjust the script accordingly.
We also check that the intervals during which precipitation water has been collected for each measurement correspond roughly to one month:

```{r GNIP_bad_rows}
rows_missing_or_unreliable_info <- is.na(rawGNIP$H2) |
                                   is.na(rawGNIP$day.span) |
                                   rawGNIP$day.span > -25 |
                                   rawGNIP$day.span < -35 
```

Third, we only keep the rows and columns we are interested in:

```{r GNIP_select}
columns_to_keep <- c("Name.of.Site", "Latitude", "Longitude", "Altitude",
                     "Year", "Month", "H2")
GNIPData <- rawGNIP[!rows_missing_or_unreliable_info, columns_to_keep]
```

Fourth, we turn the variable `Name.of.Site` into a factor:

```{r GNIP_factor}
GNIPData$Name.of.Site <- as.factor(GNIPData$Name.of.Site)
```

Last, we rename the columns to conform to the general IsoriX format and we check that the data seem correct:

```{r GNIP_change_column_names}
colnames(GNIPData) <- c("source_ID", "lat", "long", "elev", "year", "month", "source_value")
```

This is what the beginning of `GNIPData` looks like:

```{r GNIP_show, eval = FALSE}
head(GNIPData)
```

```{r GNIP_show_bis, echo = FALSE}
kable(head(GNIPData))
```

As you can see, the format is the same as the one for `GNIPDataDE`, which is what we want.


## Processing the raw data {#processing}

In order to build your isoscape with IsoriX, your dataset must be aggregated in such a way that each observation corresponds to the mean and variance of isotope values collected in one location, and potentially over fixed periods of time.

The time interval considered corresponds to the interval for which single isoscapes will be build.
For example, if the interval is the month you will end up with one set of isoscapes for each month of the year (even if those can later be aggregated to form a yearly isoscape as we shall see in section XX [to come]). 

To aggregate the data, you can choose to aggregate your dataset on your own, or to use our function `prepsources()`.
This function allows you to aggregate the data over different time intervals.
It also allows you to apply some restriction on the data, such as selecting only particular months, years, or locations.

Here we will use the function `prepsources()` to select from the dataset `GNIPData` the observations available within an extent of latitude and longitude that covers roughly Europe.

### Aggregation for a single set of isoscapes {#singleset}

We will start by using the default aggregation scheme, which simply produces an overall average per location.

```{r GNIP_EU_build}
GNIPDataEUagg <- prepsources(data = GNIPData,
                             long_min = -30, long_max = 60,
                             lat_min = 30, lat_max = 70)
```


Let us now visualize the beginning of the dataset:

```{r GNIPDataEU_view, eval = FALSE}
head(GNIPDataEUagg)
```

```{r GNIPDataEU_view_bis, echo = FALSE}
kable(head(GNIPDataEUagg)) %>% kable_styling(bootstrap_options = "striped", font_size = 15)
```


As you can see, we now have observation for each location.
The column `mean_source_value` gives the average across months and years.
The column `var_source_value` gives the variance between monthly measurements spanning across years.


### Aggregation for multiple sets of isoscapes {#multiplesets}

Depending on what you want to do, you may need to prepare the data for multiple sets of isoscapes.
This is for example the case if you want to ultimately build a yearly isoscape by attributing different weights of the data from each month, as it is the case for the widely use annual precipitation-weighted isoscape.
We will choose this specific  example and thus prepare the data in view of building 12 sets of isoscapes (one for each month).

This can be done again with the use of the function  `prepsources()` using the argument `split_by = "month"`:

```{r GNIP_EU_12_build}
GNIPDataEU12agg <- prepsources(data = GNIPData,
                               long_min = -30, long_max = 60,
                               lat_min = 30, lat_max = 70,
                               split_by = "month")
```

Let us visualize the first 15 lines of the dataset:

```{r GNIPDataEU12_view, eval = FALSE}
head(GNIPDataEU12agg, n = 15)
```

```{r GNIPDataEU12_view_bis, echo = FALSE}
kable(head(GNIPDataEU12agg, n = 15)) %>% kable_styling(bootstrap_options = "striped", font_size = 14)
```

As you can see, we now have observation for each month.
The column `mean_source_value` gives the average for each month across the years.
The column `var_source_value` gives the between-year variance in monthly measurements.


## Fitting the geostatistical models

Isoscapes are predictions stemming from statistical models.
In this section we will show you how to fit such models based on the dataset we prepared in section \@ref(processing).
We refer to *set of isoscapes* because IsoriX does not create a single isoscape predicting the spatial distribution of isotopes (i.e. the main isoscape), but it also creates other isoscapes related to the main isoscape.
For example, it also creates an isoscape indicating how reliable the predictions are in any particular locations.
If a single dataset is being used (as in section \@ref(singleset)), then a single set of isoscapes need to be built.
Instead, if several datasets have been prepared (as in section \@ref(multiplesets)), then you will have to accordingly build multiple sets of isoscapes.

### Fitting the models for a single set of isoscapes {#isofit}

To fit the geostatisical model required to build a single set of isoscapes, you need to use the function `isofit()`.
This function has several parameters that can be adjusted to fit different models (see `?isofit`).
Here we will consider two fixed effects (the elevation and the absolute latitude) on top of the default setting:

```{r isofit fake, eval = FALSE}
EuropeFit <- isofit(data = GNIPDataEUagg,
                    mean_model_fix = list(elev = TRUE, lat_abs = TRUE))
```

```{r isofit real, echo = FALSE, results = FALSE}
if (file.exists("output/EuropeFit.rda")) {
  load("output/EuropeFit.rda")
} else {
  EuropeFit <- isofit(data = GNIPDataEUagg,
                      mean_model_fix = list(elev = TRUE, lat_abs = TRUE))
}
```

### Fitting the models for a multiple set of isoscapes {#isomultifit}

If you have created multiple sets of data, you need to use the function `isomultifit()` for fitting one set of isoscapes per set of data.
As for `isofit()`, several parameters can be adjusted to fit different models, and we will use here the settings seen previously.

```{r isomultifit fake, eval = FALSE}
EuropeFit12 <- isomultifit(data = GNIPDataEU12agg,
                           mean_model_fix = list(elev = TRUE, lat_abs = TRUE))
```

```{r isomultifit real, echo = FALSE, results = FALSE}
if (file.exists("output/EuropeFit12.rda")) {
  load("output/EuropeFit12.rda")
} else {
  EuropeFit12 <- isomultifit(data = GNIPDataEU12agg,
                             mean_model_fix = list(elev = TRUE, lat_abs = TRUE))
}
```

For the time being, the function isomultifit()` fits sequentially the sets of isoscapes.
So for the example used here, it means that 12 sets of isoscapes are fitted sequentially, which thus takes more computing time than building a single set of isoscape.
In the future our plan is to revise the internal structure of this function for it to be able to use several CPUs simultaneously and thus benefit from the IsoriX option used to define the number of CPUs to work with (see section \@ref(globaloptions)).

## Saving the fitted models

Fitting models can take substantial computing time, depending on the amount of data you have and on the complexity of the models to be fit.
Therefore, you may want to save the fitted models in order to be allowed to reuse them later (even after having opened a new R session), or in order to send them to your colleagues.

This can be done as follow:

```{r saving models fake, eval = FALSE}
save(EuropeFit, file = "EuropeFit.rda", compress = "xz")
save(EuropeFit12, file = "EuropeFit12.rda", compress = "xz")
```

```{r saving models real, echo = FALSE, warning=FALSE}
if (!file.exists("output/EuropeFit.rda"))
  save(EuropeFit, file = "output/EuropeFit.rda", compress = "xz")
if (!file.exists("output/EuropeFit12.rda"))
  save(EuropeFit12, file = "output/EuropeFit12.rda", compress = "xz")
```

The function `save()` will (by default) store your R object in a file that can be found in your working directory.
To use `save()`, we must provide the object you want as a first argument of the function.
Then, `file =` defines the name of the file that will store the R object in your hard drive.
You can also include a path to this name so to store the file wherever you want.
This name can be different from the name object but naming the file as the object allows you to remember what the name of the stored object is (check `?saveRDS` for an alternative way of saving R objects which allows for you to name objects when loading them).
The last argument `compress =` is optional; it allows for the creation of smaller files without loosing any content, so we always use it.

For loading a saved object (in a new session of R for example), just use the function `load()` as follows (but make sure the saved object is in your working directory or include the path to the file names):

```{r loading models, eval = FALSE}
load(file = "EuropeFit.rda")
load(file = "EuropeFit12.rda")
```

_Be careful_, we do not recommend you to reuse saved object after updating either IsoriX, or spaMM, or both.
After one of such update, the best practice to make sure that every thing is working properly is to refit all your models.
By doing this you may also benefit from potentially new improvements we would have implemented. 

## Examining the fitted models

### Plotting basic information about the models

You can display some information about the model fits by typing:

```{r plot isofit fake, eval = FALSE}
plot(EuropeFit)
```

```{r plot_isofit_real, echo = FALSE, out.width = '80%', fig.show = 'hold', fig.align = 'center', fig.width = 6, fig.height = 6}
old_opt <- options_IsoriX(dont_ask = TRUE)
plot(EuropeFit)
options_IsoriX(old_opt)
```

In the panel produced, the left column shows the relationship between the observed and predicted response (top) and the variation in spatial autocorrelation with the distance between location (bottom) captured by the model for the fit called `mean_fit`, which corresponds to the fit of the mean isotopic values.
The right column shows the same information for the fit called `disp_fit`, which corresponds to the fit of the residual dispersion variance in the isotope values.
On the first row you can see points distributed practically along the 1:1 diagonal.
A different slope would suggest a high residual variance of the data.
We do not expect the points to fall exactly on the line because the model fit does not attempt to predict perfectly the observations used during the fit.
Instead, the model fit produces a smooth surface that aims at reaching a maximum predictive power for locations not considered during the fit.
The second row gives you an idea of the strength of the spatial autocorrelation captured by the models.
Here the results suggest that the autocorrelation is very strong.
Not considering this autocorrelation would thus lead here to a poor approximation of the isoscape and resulting assignments.


### Examining the summary tables

To simply explore the fitted models you can simply type their name and the information of each fit will be displayed:

```{r summary isofit}
EuropeFit
```

The object `EuropeFit` created in section \@ref(isofit) is a list containing the two fits aforementioned: `mean_fit` and `disp_fit`.
This is why you see information about those two corresponding fits.
These summary tables are created behind the scenes by [**spaMM**](https://kimura.univ-montp2.fr/~rousset/spaMM.htm), so, learning a little about spaMM and mixed models is useful to derive meanings from these outputs.
(We are also planning to write an extensive description of model outputs to guide you.)


### Doing more stuff with the fitted models

Each of model fit is an object of class `HLfit` which has been created by spaMM.
You can thus directly manipulate those fits using spaMM for advance usage.
For example, if you want to compare different models using information criterion, you can call the `AIC()` function implemented by spaMM by simply typing:

```{r AIC EuropeFit, eval = FALSE}
AIC(EuropeFit$mean_fit)
```

```{r print AIC EuropeFit, echo = FALSE}
print(AIC(EuropeFit$mean_fit))
```

Note that we re-exported some spaMM functions for you to use without the need to load the package spaMM (e.g. `AIC.HLfit()` which is implicitly here), but you will have to call `library(spaMM)` to access all spaMM functions.


## Building the isoscapes

### Prepare the structural raster

To build the isoscape using the geostatistical models fitted above, you need to provide the function `isoscape()` with a structural raster.
Such raster defines the locations at which we want to predict the isotopic values.
It must also contains the values for the predictors used in the geostatistical models at such locations.

We will thus start by preparing such structural raster.
Here, as a basis for the structural raster, we will download a high resolution elevation raster from Internet using our function `getelev`:

```{r getelev fake, eval = FALSE}
getelev(file = "input/elevation_world_z5.tif")
```

```{r getelev real, eval = TRUE, echo = FALSE}
if (!file.exists("input/elevation_world_z5.tif")) {
  getelev(file = "input/elevation_world_z5.tif")
}
```

You may not need to pass any argument to `getelev()`, but here we opted to choose to define ourselves where to store the elevation raster and how to call such file using `file = "input/elevation_EU_z5.tif"`.
We did not alter the resolution which is high enough under the default setting for our example, but doing so would be easy: one simply needs to define a value lower or larger than 5 for an argument called `z` (see `?getelev()` for details and [here](https://github.com/tilezen/joerd/blob/master/docs/data-sources.md#what-is-the-ground-resolution) for information on what `z` really implies).

We then import the high resolution elevation raster and transform it into an R object of class `RasterLayer` object using the package raster:

```{r elevationraster, dev='CairoPNG'}
## we import the high resolution raster
ElevWorld <- raster("input/elevation_world_z5.tif")

## we check that the import worked
ElevWorld

## we draw a simple plot to check what we just downloaded
plot(ElevWorld)
```

Then, you may need to *crop* and *resize* this raster to the appropriate size before creating the isoscapes.

Cropping is used to define a rectangular area over which we will predict the isotope values.
If the area you define to build your isoscapes is too small, you may miss crucial information.
For example, a set of isoscapes that cover an area that is too small may lead you to miss the real origin of your migratory organisms during the assignment step.
Similarly, if the area to be used for building the isoscapes is too large, you may extrapolate and the model will thus create unreliable predictions.

Resizing the raster is used to reduce its resolution by aggregating neighboring cells in the raster.
If the selected resolution is too large, computations may take too much time or memory to work.
If the selected resolution is too coarse, you may smooth too much the predictors over the landscape (e.g. the peaks and valleys of a mountain chain will be reduced to a plateau), which will lead to a unrealistic isoscapes that are too smooth.
To crop and resize the structural raster, you can use our function `prepraster()`.

::: {.rmdnote}
**Geeky note**: In principle you could directly crop and resize the raster to the dimensions you want before downloading it (`getelev()` does allows for you to do that).
However, in practice, we recommend you to download a raster that is a larger than what you need (`getelev()` downloads the full world by default, which should always be sufficient) and with a resolution overshooting a little what you may need.
The benefit of relying on `prepraster()` rather than `getelev()` to crop and downsize the structural raster is that if you don't like the result, you can simply rerun `prepraster()` again -- with different settings -- without having to download again a large amount of data from the net.
The only reason to rerun `getelev()` would be if you think the resolution of your initial raster is too low.
:::

Here, the structural raster downloaded with `getelev()` covers the entire planet, so we will crop it to the limit of Europe which we also used to select our source data (see section \@ref(processing)).

To crop the elevation raster to a particular size, you can either provide a set of coordinates to `prepraster()`by means of the argument `manual_crop`.
Alternatively, you can simply provide a fitted model to the function `prepraster()` by means of the argument `isofit`.
In this latter case, the function will determine the coordinates required for the cropping automatically, from the coordinates of the weather stations used during the fitting procedure.
We choose here to use this simpler way of defining the cropping area:

```{r build ElevEurope fake, eval = FALSE}
ElevEurope <- prepraster(raster = ElevWorld,
                         isofit = EuropeFit,
                         aggregation_factor = 4)
```

```{r build ElevEurope real, echo = FALSE, results = FALSE}
if (file.exists("output/ElevEurope.rda")) {
  load("output/ElevEurope.rda") } else {
  ElevEurope <- prepraster(raster = ElevWorld,
                           isofit = EuropeFit,
                           aggregation_factor = 4)
  if (raster::inMemory(ElevEurope)) {
    save(ElevEurope, file = "output/ElevEurope.rda", compress = "xz")
  }
}
```

You may see that we also chose here to reduce the resolution of the elevation raster by choosing an aggregation factor of 4.
We recommend to first use a large aggregation factor (e.g. 10) to draft your workflow and quickly notice errors in your code.
Then, once things are working as they should, we recommend you to use the lowest aggregation factor that you hardware and patience can handle (i.e. 1 if possible).
As mentioned above, if you need isoscapes with a even greater resolution, use `getelev()` with a value of the parameter `z` that is higher than 5 (an increment of 1 -- i.e. `z = 6` already makes a big difference as shown [here](https://github.com/tilezen/joerd/blob/master/docs/data-sources.md#what-is-the-ground-resolution)).
Here we set the `aggregation_factor` to 4 as something in between which produces plots nice enough for this tutorial, without being too heavy for us to recompile this bookdown quickly; but you should use a higher resolution than used here for real applications.

You can easily check what your structural raster looks like by plotting it.
Compared to the map plotting above for `ElevWorld`, we will now draw a slightly more advanced map using the functions of the packages [**lattice**](http://lattice.r-forge.r-project.org) and [**rasterVis**](https://oscarperpinan.github.io/rastervis/) which we made available in IsoriX:

```{r plot_ElevEurope, dev='CairoPNG'}
levelplot(ElevEurope,
          margin = FALSE,
          col.regions = viridisLite::turbo, ## the colour palette
          main = "Structural raster") + 
layer(sp.polygons(CountryBorders, col = "white")) +
layer(sp.polygons(OceanMask, col = "white", fill = "lightgreen"))
```

Note that, in this raster, the highest elevation does not correspond to the highest elevations that are known in the depicted area (the maximal value in the raster is `r round(max(values(ElevEurope)))`m, but Mount Elbrus is 5642m tall).
This is not a bug but an expected result of aggregation.
Indeed, each cell of the raster only contains a single value which corresponds to a summary statistic (the mean) applied to all locations aggregated together in the cell.

::: {.rmdnote}
**Geeky note**: You may change what summary statistic is applied when using `prepraster()` using the argument `aggregation_fn`, but you have no control over the aggregation happening within `getelev()` since it uses ultimately a package that does not allow for controlling the summary statistics.
Whatever you do, think carefully on the consequences of your choice upon how you are planning to use the isoscapes (e.g. for assigning species living only at the top of the mountains, using `aggregation_fn = max` may make more sense than the default).
:::

### Predicting the isoscapes

We will now build the isoscapes using the function `isoscape()`.
The function simply takes the structural raster and the fitted model as arguments:

```{r build EuropeIsoscape fake, eval = FALSE}
EuropeIsoscape <- isoscape(raster = ElevEurope,
                           isofit = EuropeFit)
```

```{r build EuropeIsoscape real, warning = FALSE, echo = FALSE, results = FALSE}
if (file.exists("output/EuropeIsoscape.rda")) {
  load("output/EuropeIsoscape.rda")
} else {
  EuropeIsoscape <- isoscape(raster = ElevEurope,
                             isofit = EuropeFit)
  if (raster::inMemory(EuropeIsoscape$isoscapes)) {
    save(EuropeIsoscape, file = "output/EuropeIsoscape.rda", compress = "xz")
  }
}
```

The function `isoscape()` creates 8 different rasters that correspond to different aspects of the isoscape (see `?isoscape()` for details).

We can check which rasters have been created by typing:

```{r show EuropeIsoscape}
EuropeIsoscape$isoscapes
```


### More complex isoscapes? The example of building the precipitation-weighted annual average isoscapes

[to come]


## Plotting the isoscapes

One particular strength of IsoriX is the ability to produce nice looking and highly customisable plots of the isoscapes and assignment maps.
All plotting functions in IsoriX present default settings so that a simple call to `plot()` on your object is sufficient to generate a nice looking plot.

For example, here you can obtain the main isoscape by typing:

```{r plot_EuropeIsoscape, dev='CairoPNG'}
plot(EuropeIsoscape)
```


This shows the prediction of the average isotope value in space (i.e. the point predictions).


### Other isoscapes

You can also plot any of the other isoscapes by specifying the name of the raster using the argument `which`.
As examples, we are now going to plot the residual variation in isotope values predicted at each location, and the prediction variance:

```{r plot_residVar, dev='CairoPNG'}
plot(EuropeIsoscape, which = "mean_residVar")
```

This first isoscape quantifies the temporal variation at each location.
Here the variation represents the between month variance spanning across years, but it will not always be so as it depends on what how you aggregated the data (see section \@ref(processing)).
Interestingly, you can see that the variance is large and spatially structured.
Believe or not, but most methods of assignment alternative to IsoriX consider such variance to be equal to zero!

```{r plot_predVar, dev='CairoPNG'}
plot(EuropeIsoscape, which = "mean_predVar")
```

This second isoscape quantifies the uncertainty in your point predictions.
As you can see, the uncertainty increases when you go far from the source location.
This isoscape can help to update the design of the data collection.
It is also used internally when performing assignment.
Indeed, if you are very uncertain of what the average isotope value is in a location, then you cannot rule out that any organism may have come from that location.
It is unfortunate but an inescapable reality.


### Fiddling with the plots {#plotisoscapes}

If you want to change anything on the existing plot, have a look at `?plots`.
The help file should provide you with all the details you need to customize your plots.
Here is are a couples of example:

```{r plot_EuropeIsoscape2, dev='CairoPNG'}
plot(EuropeIsoscape,
     sources = list(pch = 3, col = "orange"),
     borders = list(col = "white"),
     mask    = list(fill = "darkgrey"),
     palette = list(range = c(-130, 10), step = 1, n_labels = 10, fn = "rainbow"))
```

```{r plot_EuropeIsoscape3, dev='CairoPNG'}
plot(EuropeIsoscape,
     title = "H Isoscape",
     sources = list(draw = FALSE),
     borders = list(borders = NULL),
     mask    = list(fill = "black"),
     palette = list(range = c(-130, 20), step = 30, fn = NULL))  
```

For example, you can see that it is possible to provide a function defining the colours using the argument `fn` from the list `palette`.
Note that when it is set to `NULL`, the famous palette [**viridis**](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html) is used instead of our default palette.

If you want to add things on top of the plots, you can do it as you would do it for any plot created using the R graphical package [**lattice**](http://lattice.r-forge.r-project.org/index.php).
The package lattice is very powerful but not always very easy to use.
We may also provide you with an implementation of IsoriX around the now widely used package [**ggplot2**](https://ggplot2.tidyverse.org) but that would imply a lot of work so we are still thinking about it...

We will also add plotting tutorials to this bookdown.


### Why you should save plots or how to export nice looking plots? {#Cairo}

Displaying the plot directly using R is very inefficient when it comes to high resolution isoscapes.
It may take a very long time and it may not fully work.

It is better to directly plot the figures into files.

::: {.rmdnote}
**Geeky note**: Since the lattice system is not part of ggplot2, the function `ggsave()` should not be used.
:::


#### Using base R graphics devices functions

A first option is to save your plot using one of the base R function such as `png()` or `tiff()`.


For example, you could do:

```{r png, results = FALSE, eval = FALSE, results = FALSE}
png(filename = "output/Myisoscape.png",
     width = 1920,
     height = 1080,
     res = 200)
plot(EuropeIsoscape)
dev.off()
```

or

```{r tiff, results = FALSE, eval = FALSE, results = FALSE}
tiff(filename = "output/Myisoscape.tiff",
     width = 1920,
     height = 1080,
     res = 200)
plot(EuropeIsoscape)
dev.off()
```

As you can see, you first initialize the creation of the plot with the function `png()` or `tiff()`, then you call your plot, then you tell your computer that you are done with `dev.off()`.

As arguments, you probably want to specify the dimensions and the resolution of the file.
The height and width are here considered to be in pixels (default setting, you can choose other units if you want using the argument `units`). Here we chose the so-called Full-HD or 1080p resolution (1080x1920) because we wanted to observe the isoscape carefully on a monitor of that resolution.
If your screen is Full-HD, try it and display the plot in full screen to get better results (if the plot does not match the definition of your screen, things can get ugly), if your screen is not, try another resolution.
If you don't know what resolution your screen has, you can visit http://www.whatismyscreenresolution.com/.

The parameter `res` is very useful as it rescales the line and fonts in the plot.
So if everything is too small just increase the value and if everything looks too bold and ugly, decreases it.

You also should specify the `filename`.
If like us, you don't indicate the path in the `filename`, the file will be stored in your working directory which you can easily see by typing `getwd()`.
So after running such code, simply go in your working directory and open the right file.

We prefer the PNGs files since they are much lighter than their TIFF counterpart, but one format may lead to a better output for you.
The nice thing about these functions is that there do not require you to install anything extra.
They also allow you to set the resolution, which is sometimes asked by journals.
So if that works for you do stick to that.

Unfortunately, depending on your system, usual plotting functions provided with R may not always work well for high resolution rasters.
For example, using `png()` or `tiff()` may lead to display artifacts showing some random white lines.
Changing the value of the argument `res` is often enough to get rid of these lines.
Another more effective way is to use the package Cairo.


#### Using Cairo graphics devices functions

One set of plotting functions that seem to never suffer from white line artifacts are those provided by the package [**Cairo**](http://www.rforge.net/Cairo/).

To use such functions, you first must make sure Cairo is installed and can be loaded:

```{r load Cairo}
library(Cairo)
```

If it complains, try to see why.
If you have not installed it, just install the package.
One noticeable source of trouble comes with MacOS.
On this operating system, the package Cairo does not always install/load successfully.
Problems happen when the program [**xquartz**](https://www.xquartz.org/) is not present on the system.
This program used to be shipped by defaults on old version of MacOS but it is no longer the case.
So if you are in trouble, please start by installing xquartz (outside R).

If Cairo is installed and loaded successfully, typing `?Cairo` will show you all possible file formats you can save your plots into (PNG, JPG, TIFF, PDF...).

Here we will show how to save our isoscape both as a PNG and as a PDF.

Let us start by creating a PNG file with the main isoscape:

```{r Cairo PNG, results = FALSE, eval = FALSE, results = FALSE}
CairoPNG(filename = "output/Myisoscape.png",
         height = 1080,
         width = 1920,
         res = 200)
plot(EuropeIsoscape)
dev.off()
```

As you can see, once you have managed to install Cairo it becomes as simple as using the base R functions.

PDFs derived from raster objects tend to be heavy and to be badly rendered by most viewing software, so we recommend you to stick to PNGs.
Yet, if you must, here is how to create PDFs using Cairo:

```{r Cairo PDF, results = FALSE, eval = FALSE, results = FALSE}
CairoPDF(file = "output/Myisoscape.pdf",
         height = 10,
         width = 15)
plot(EuropeIsoscape)
dev.off()
```

The arguments slightly differ with the function creating PNGs.
The argument setting the name of the file is now called `file` and not `filename`, the resolution has now to be provided in inches and the argument `res` does not exist for PDFs, but default results are usually fine.

## Next?

If your goal was to produce isoscapes, you should be done by now.
The following steps of this tutorial are for users interested in inferring the origin of organisms based on their isotopic signature.
