
### Changing font size? the example of the IsoriX welcome picture

You can influence the font size while exporting your plots. Here is for example how we created the welcome picture of the __IsoriX__ page on GitHub (\url{www.github.com/courtiol/isorix_project}):

```{r welcome picture, fig.show="hide", eval=do.eval}
plot.welcome.isoscape <- plot(x = isoscape,
                              palette = list(n.labels = 15,
                                             range = c(-130, 10),
                                             digits = 1))

CairoPNG(file = "./output/isoscape.png",
         units = "in",
         width = 10,
         height = 8,
         pointsize = 24*92/72,
         dpi = 96)

plot.welcome.isoscape 

dev.off()
```


## Saving IsoriX plots without Cairo

If you really don't manage to use __Cairo__ or if you don't want to use it, you can always use the simple out-of-the-box alternative offered by R. You just need to use different function calls:

* `CairoPNG()` --> `png()`

* `CairoPDF()` --> `pdf()`

Nothing else changes, the inputs should be the same and you still need to close the graphic device using `dev.off()` after ploting.




## The End

That is all for now! Here are the information of the R session we used:

```{r R session}
sessionInfo()
```



