# Contributing to IsoriX

In this chapter we will give you tips on how you can help us to make _IsoriX_ better.

[more to come]

## Coding conventions

This section is only intended to those who would like to contribute source code.
It describes the conventions one should follow while coding for _IsoriX_.
Note that we have not always followed such conventions but we will progressivelly reformat the code in such way.

### Object assignment

Object assignments are done using an arrow `<-`, not `=`.
```{r object assignment, eval = FALSE}
## Example:
y <- 2
```

### Notation and naming

* datasets should be named in the format `SomethingSomething`.

* classes should have a name of the format `SOMETHING`.

* internal functions should have a name of the format `dosomething_somethingelse`, where the input is `something_somethingelse` and the output for the global environment should be `my_something`.

### Curly braces

The opening curly braces are separated with one white space and the closing ones go on a new line and are aligned with the first command.
The else statement is surrounded by curly braces on the same line.

```{r curly, eval = FALSE}
## Example for an if-condition:
if (y == 0) {
  log(y)
} else {
  y ^ 2
}
```

The same goes for a function except we do not put a space before the left parentheses.

```{r eval = FALSE}
## Example for a function:
function(x) {
  x + x
}
```

### Indentation and spacing

We have 2 spaces indentation throughout the whole code.
An exception is made when a line break occurs inside parentheses, then the wrapped line is aligned with the first character inside the parenthesis.
```{r indentation, eval = FALSE}
## Example:
plot(x = sort(rnorm(47)),
     type = "s",
     main = "title")
```

Spaces go around all operators (```+```, ```=```, ```<-```, ```!=```, ```==```, etc.) for improved readability.

### Commenting guidelines

For comments providing explanation inside the code we use a double hash ```##``` in order to distinguish such comments from commented code which we comment using a single hash ```#```.

[more to come]
