---
title: "How to export your IsoriX objetcs to GIS?"
author: "Alexandre Courtiol, François Rousset & Stephanie Kramer-Schadt"
date: "`r Sys.Date()`"
output:
    pdf_document
vignette: >
  %\VignetteIndexEntry{Export to GIS in IsoriX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, initialization, echo=FALSE}
library(knitr)
set.seed(123)  ## set seed for entire manuscript
# set global options for knitr (if changed, all cashed chunked will be rerun!)
knitr::opts_chunk$set(cache=FALSE)
do.eval <- FALSE ## to check that it does work, put TRUE
```

Welcome to __IsoriX__, in this vignette we present the steps required for you to export your isoscape or assignment maps for using them outside R within the GIS program of your choice.

## Before starting

Please read the vignette ___Workflow___, if you haven't done so. You can access it by simply typing:

```{r, workflow, eval=FALSE}
vignette("Workflow", package="IsoriX")
```

We assume that you are still in the same R session than the one created during the simple workflow example. That implies that __IsoriX__ is loaded and that the objects we will use below have already been created.


```{r, load isoscape, echo=FALSE, message=FALSE}
library(IsoriX)
load("../vignette_workflow/isoscape.rda")
```

## Export your data to a GIS
R comes along with a suite of packages designed for easy handling of spatial data and exchange formats with standard GIS programs, such asArcGIS by ESRI. In the following, we will first explain the different spatial data formats stored within the objects created in the IsoriX-package and explain how they can easily be exported. We then provide some visualisation functions along with coordinate reference system transformations to directly handle the spatial layers in R.

## RasterLayers, RasterStacks (RasterBricks),  and SpatialPolygons, SpatialPointsDataFrames
There are four types of spatial objects in IsoriX that one should know. RasterLayers are 2-dimensional matrix-like grids, with each grid cell containing a value representing information, e.g. the mean predicted isotopic value or elevation. RasterStacks or RasterBricks are multilayer-raster objects, i.e. multi-band satellite images (see __raster__ package).
SpatialPolygons and SpatialPoints are vector-data, i.e. geometric objects of polygons, (lines) or points and can be handled through the __sp__ package.

## Accessing the spatial information stored in the Isoscape- and Isorix-objects
The 'worldcountries' and the 'oceanmask' are ready-to-use polygon vector layers stored in IsoriX as SpatialPolygons. Further spatial information is stored in the Isoscape- and the Isorix-objects.

### The Isoscape-object
When you type
```{r AccessIsoscape1}
unlist(isoscape) #note: isoscape = the Isoscape-object
```
you will see that there is information stored in $Isoscape (RasterStack) and $sp.points (SpatialPointsDataFrame). The RasterStack consists of 8 layers specified in the names-slot. Typing
```{r AccessIsoscape2}
names(isoscape$Isoscape) 
```
shows you which RasterLayers are stored in the stack. To access a single RasterLayers, e.g. to plot or export it, type
```{r AccessIsoscape3}
isoscape$Isoscape[['mean']] # similar to: isoscape$Isoscape[[1]]
```
In a similar way, you can access the weather station data, called sources:
```{r AccessIsoscape4}
isoscape$sp.points  
```

### The Isorix-object
Analoguously, one can access all the information stored in the Isorix-object, that is, the assignment. Here, for each statistical output (e.g. p-value), a RasterBrick is stored containing the information for each single individual entered into the assignment dataset.
```{r AccessIsorix1}
unlist(assignment)
```

That is, the assignment probability of e.g. a given individual 'Mbe_1' can be extracted from the RasterBrick $indiv.pv in the following way:
```{r AccessIsorix2}
assignment$indiv$pv[['Mbe_1']]
```

The Isorix-object also stores the sources (e.g. water stations used to create the isoscape) as well as the locations of the calibrationdata as SpatialPointsDataFrames and can be accessed accordingly:

```{r AccessIsorix3}
assignment$sp.points$sources
assignment$sp.points$calibs
```


## Export spatial objects created in IsoriX 
The ESRI shapefile is a widely used interchange format to store vector data, i.e. in our case the SpatialPolygons (worldcountries, oceanmask) and the SpatialPointsDataFrames (sources, calibration locations). There are two libraries in R that can be used: rgdal and maptools (https://www.nceas.ucsb.edu/scicomp/usecases/ReadWriteESRIShapeFiles). To save SpatialPolygons, a little workaround is needed to join a dataframe (attribute table) to each feature. Before the export, make sure that the projection is defined for all spatial data. By default, we are working with WGS84.


```{r ExportShape,eval=FALSE,tidy.opts=list(keep.blank.line=FALSE, width.cutoff=60)}
# the sources
SourceLocations <- assignment$sp.points$sources
projection(SourceLocations)

#workaround for worldcountries
df <- data.frame(id = getSpPPolygonsIDSlots(worldcountries)) #add attribute table with 'id' for 1625 features
row.names(df) <- getSpPPolygonsIDSlots(worldcountries)
wrld_cntrs <- SpatialPolygonsDataFrame(worldcountries, data =df)
projection(wrld_cntrs)

# export
setwd('C:/Temp') #or wherever on your harddrive
library(rgdal)
writeOGR(obj=SourceLocations, dsn=getwd(), layer='SourceLocs_rgdal', driver ='ESRI Shapefile',overwrite=TRUE)
writeOGR(obj=wrld_cntrs, dsn=getwd(), layer='SourceLocs_rgdal', driver ='ESRI Shapefile',overwrite=TRUE)

library(maptools)
writePointsShape(x=SourceLocations, "SourceLocs_mapt") #points
writePolyShape(x=wrld_cntrs, "worldcntrs_mapt") #polygons

```

RasterLayers can be exported with the __raster__ package as ascii format or geotiff. More raster formats are available, depending on the GIS-system you have. e.g. SAGA, IDRISI or Erdas Imagine (type ?writeRaster)
```{r ExportRaster,eval=FALSE,tidy.opts=list(keep.blank.line=FALSE, width.cutoff=60)}
Assignment_Prob_Mbe_1 <- assignment$indiv$pv[['Mbe_1']]
projection(Assignment_Prob_Mbe_1)

writeRaster(Assignment_Prob_Mbe_1, filename="ap1.asc",format="ascii",overwrite=TRUE,NAflag = -9999)
writeRaster(Assignment_Prob_Mbe_1, filename="ap2.tif",format="GTiff",overwrite=TRUE,NAflag = -9999)#requires rgdal
```



## Pimping up your plots
### Add a 3D-image to your plot
```{r hillshade I}
slope <- terrain(x = elev, opt = "slope",unit = "radians", neighbors = 8)
aspect <- terrain(x = elev, opt = "aspect",unit = "radians", neighbors = 8)
hillsh <- hillShade(slope, aspect, angle = 45,direction = 270)
```

```{r hillshade II,warning=FALSE,fig.pos='t', fig.width=4, fig.height=3}
plot(hillsh, col=grey(0:100/100), legend=FALSE)
#and colors on top: alpha value gives semi-transparency
image(isoscape$Isoscape[['mean']], col=terrain.colors(100,alpha=0.3),add=T)
points(isoscape$sp.points$sources,cex=0.5,pch='+')
plot(oceanmask,col='grey',border='transparent',add=T)
plot(worldcountries,col='transparent',border='white',add=T)
box()
```


### Add Gridlines
```{r gridline, fig.pos='t', fig.width=3, fig.height=3}

grd <- gridlines(isoscape$Isoscape[['mean']], 
                    easts = pretty(bbox(isoscape$Isoscape[['mean']])[1,]),
                    norths = pretty(bbox(isoscape$Isoscape[['mean']])[2,]), 
                    ndiscr = 20)
plot(grd, add=T, col='pink', lty=2)
```


### Add North Arrow and Scale Bar
```{r north arrow, fig.pos='t', fig.width=3, fig.height=3}
library(GISTools)
library(maps)
north.arrow(-20, 40, 2.25, cex.lab=0.7, col="grey80")
map.scale(-15, 35, 10,"Km",1, 1000, sfcol='red')   #should be approx correct
```


### Change the projection of your map
For certain parts of the world it is better to use a different projection than our geographic coordinate system (WGS84). Data can easily be projected in R, however, there is no 'on the fly' projection like in GIS systems - each layer has to be projected separately. To find the optimal reference system for your area, please check here:
http://www.spatialreference.org
As an example, we will use a Mollenweide-projection. There is different commands for vector and raster data.

```{r projections}
sources_moll <-  spTransform(isoscape$sp.points$sources,   CRS("+proj=moll +datum=WGS84"))
wrld_moll <- spTransform(worldcountries,   CRS("+proj=moll +datum=WGS84"))
ocma_moll <- spTransform(oceanmask,   CRS("+proj=moll +datum=WGS84")) ##### ALEX: the oceanmask get inverted, shiott...

iso_mean_moll <- projectRaster(isoscape$Isoscape[['mean']], crs=CRS("+proj=moll +datum=WGS84"))
hillsh_moll <- projectRaster(hillsh, crs=CRS("+proj=moll +datum=WGS84"))
```


### Redo the plot

```{r hillshade IIb,warning=FALSE,fig.pos='t', fig.width=4, fig.height=3}
plot(hillsh_moll, col=grey(0:100/100), legend=FALSE)
image(iso_mean_moll, col=terrain.colors(100,alpha=0.3),add=T)
points(sources_moll,cex=0.5,pch='+')
#plot(ocma_moll,col='grey',border='transparent',add=T)############## taken out
plot(wrld_moll,col='transparent',border='white',add=T)
box()

```







